---
title: "ADA Final Project"
author: "Keilya Kasprzyk, Lucy Ingaiza, Trisha Grewal"
date: "4/29/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, results = 'hide', warning = FALSE, message = FALSE)
```


```{r}
#install packages if needed

#install.packages("pacman")
#install.packages("nhanesA")
#install.packages("tidyverse")
#install.packages("car")
#install.packages("table1")
#install.packages("ggpubr")
#install.packages("gmodels")
#install.packages("nnet")
#install.packages("MASS")
#install.packages("lmtest")
#install.packages("sjPlot")
#install.packages("sjmisc")
#install.packages("sjlabelled")

#Load libraries
pacman::p_load(nhanesA, tidyverse, car, table1, ggpubr, gmodels, nnet, MASS, lmtest, sjPlot, sjmisc, sjlabelled)
```

```{r}
#Load the data and create dataset

#get questionnaire data from 2017 NHANES
nhanes2017  <- nhanesTables('Q', 2017, namesonly=TRUE)
nhanes2017tables <- lapply(nhanes2017, nhanes)
names(nhanes2017tables) <- nhanes2017

#get demographic data from 2017 NHANES
nhanesdemo2017  <- nhanesTables('DEMO', 2017, namesonly=TRUE)
nhanesdemo2017tables <- lapply(nhanesdemo2017, nhanes)
names(nhanesdemo2017tables) <- nhanesdemo2017

#get examination data from 2017 NHANES
nhanesoral2017  <- nhanesTables('EXAM', 2017, namesonly=TRUE)
nhanesoral2017tables <- lapply(nhanesoral2017, nhanes)
names(nhanesoral2017tables) <- nhanesoral2017

#extract questionnaire about work
work <- nhanes2017tables$OCQ_J

#extract questionnaire about insurance
insurance <- nhanes2017tables$HIQ_J

#extract demographics data
demographics <- nhanesdemo2017tables$DEMO_J

#extract oral examination data
oral <- nhanesoral2017tables$OHXREF_J

#merge data
merge1 <- merge(work, insurance, by="SEQN")

merge2 <- merge(demographics, oral, by="SEQN")

merged.NHANES <- merge(merge1, merge2, by="SEQN")

#subset only desired variables
NHANES.small = subset(merged.NHANES, select = c(SEQN, RIDAGEYR, RIDRETH1, DMDCITZN, RIAGENDR, OHAREC, HIQ011, OCQ670, OCQ260, OCD270))
summary (object=NHANES.small)
```

```{r}
#Check variables and their types
sapply(NHANES.small,class)
```

```{r}
#Recode to categorical or continous 
NHANES.small.clean<-NHANES.small %>% 
  mutate(SEQN= as.numeric(SEQN)) %>% 
  mutate(RIDAGEYR= as.numeric(RIDAGEYR)) %>% 
  mutate(RIDRETH1= as.factor(RIDRETH1)) %>% 
  mutate(DMDCITZN= as.factor(DMDCITZN)) %>% 
  mutate(RIAGENDR= as.factor(RIAGENDR)) %>% 
  mutate(OHAREC=as.factor(OHAREC)) %>% 
  mutate(HIQ011= as.factor(HIQ011)) %>% 
  mutate(OCQ670= as.factor(OCQ670)) %>% 
  mutate(OCQ260=as.factor(OCQ260)) %>% 
  mutate(OCD270= as.numeric(OCD270))

```

```{r}
 #Check data types 
sapply(NHANES.small.clean,class)

```

```{r}
# Rename variables  
colnames(NHANES.small.clean)= c("Respondent id","Age","Race","Citizenship",
                                "Sex","Oral.health.status","Insurance","Schedule",
                                "Job.type","Months.working")
summary(NHANES.small.clean)
```
 

```{r}
#Recode variables
#Assign NA to don't know and refused responses
NHANES.small.clean$Citizenship<- car::recode(NHANES.small.clean$Citizenship,
                              "1 = 'US Citizen';
                               2 = 'Not a US citizen';
                               7 = NA;
                               9 = NA")
NHANES.small.clean$Insurance <- car::recode(NHANES.small.clean$Insurance,
                              "1 = 'Yes';
                               2 = 'No';
                               7 = NA;
                               9 = NA")
NHANES.small.clean$Schedule <- car::recode(NHANES.small.clean$Schedule,
                              "1 = '9am-5pm';
                               2 = 'Evenings';
                               3 = 'Early mornings';
                               5 = 'Variable';
                               7 = NA;
                               9 = NA")
NHANES.small.clean$Months.working <- car::recode(NHANES.small.clean$Months.working,
                              "77777 = NA;
                               99999 = NA")
NHANES.small.clean$Race <- car::recode(NHANES.small.clean$Race,
                              "1 = 'Mexican American';
                               2 = 'Other Hispanic';
                               3 = 'Non-Hispanic White';
                               4 = 'Non-Hispanic Black';
                               5 = 'Other Race'")
NHANES.small.clean$Oral.health.status <- car::recode(NHANES.small.clean$Oral.health.status,
                              "1 = 'Poor';
                               2 = 'Poor';
                               3 = 'Moderate';
                               4 = 'Good'")
NHANES.small.clean$Job.type <- car::recode(NHANES.small.clean$Job.type,
                              "1 = 'Private';
                               2 = 'Federal';
                               3 = 'State';
                               4 = 'Local';
                               5 = 'Self-employed';
                               6 = 'Family work';
                               77 = NA;
                               99 = NA")
NHANES.small.clean$Sex <- car::recode(NHANES.small.clean$Sex,
                              "1 = 'Male';
                               2 = 'Female'")

summary(NHANES.small.clean)

```

```{r}
# drop NAs
NHANES.complete <- NHANES.small.clean %>% 
  drop_na()

summary(NHANES.complete)
```
### Examine demographic distributions

```{r}
#Age
ggplot(data = NHANES.complete, aes(x = Age)) + 
  geom_histogram(binwidth = 5, color = "black", fill= "#8856a7") +
  xlab("Age") +
  ylab("Frequency") +
  ggtitle("Distribution of age across oral health status")

#Age boxplot
ggplot(data = NHANES.complete, aes(x = Oral.health.status, y = Age)) +
  geom_boxplot(aes(fill = Oral.health.status), alpha = .4) +
  labs(x = "Oral Health Status", y = "Age",
       title = "Distribution of age across oral health status") +
        scale_fill_discrete(name = "Oral Health Status") +
  theme_bw()

#Race
ggplot(data = NHANES.complete, aes(x = Race)) + 
  geom_bar(aes(fill = Race)) +
  labs(title = "Bar graph of race across oral health status") +
  ylab("Frequency") +
  scale_fill_manual(values=c("#8dd3c7", "#ffffb3", "#bebada", "#fb8072", "#80b1d3"), 
                    name = "Race") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

#Citizenship
ggplot(data = NHANES.complete, aes(x = Citizenship)) + 
  geom_bar(aes(fill = Citizenship)) + 
  labs(title = "Bar graph of citizenship Status across oral health status") + 
  ylab("Frequency") +
  scale_fill_manual(values=c("#7570b3", "#1b9e77"), 
                    name = "Citizenship Status") +
  theme_bw()

#Sex
ggplot(data = NHANES.complete, aes(x = Sex)) + 
  geom_bar(aes(fill = Sex)) + 
  labs(title = "Bar graph of sex across oral health status") +
  ylab("Frequency") +
  scale_fill_manual(values=c("#a6cee3", "#1f78b4"), 
                    name = "Sex") +
  theme_bw()

#Insurance
ggplot(data = NHANES.complete, aes(x = Insurance)) + 
  geom_bar(aes(fill = Insurance)) +
  labs(title = "Bar graph of insurance coverage across oral health status") +
  ylab("Frequency") +
  scale_fill_manual(values=c("#af8dc3", "#7fbf7b"), 
                    name = "Insurance Coverage") +
  theme_bw()
```

### Examine RQ predictors vs outcome

```{r}
#look at cross tabs

#work type vs oral health status
CrossTable(NHANES.complete$Job.type, NHANES.complete$Oral.health.status)

#work schedule vs oral health status
CrossTable(NHANES.complete$Schedule, NHANES.complete$Oral.health.status)

#look at distribution of years working
ggplot(data = NHANES.complete, aes(x = Months.working)) + 
  geom_histogram(binwidth = 10, color = "black", fill= "#a6cee3") +
  xlab("Months working in current job") +
  ylab("Frequency") +
  ggtitle("Distribution of months working")

#some small cells and a very skewed distribution for months working
#will recode
```

```{r}
#recode predictor variables

#job type collapse to 3 cats
NHANES.complete$Job.Description <- car::recode(NHANES.complete$Job.type, 
                              "'Federal' = 'Other';
                               'State' = 'Other';
                               'Local' = 'Other';
                               'Self-employed' = 'Other';
                               'Family work' = 'Other'")

#Make months working into categories for years in job
NHANES.complete$Years.working[NHANES.complete$Months.working <= 12]<- 1 #Less than a year
NHANES.complete$Years.working[NHANES.complete$Months.working > 12 
                                 & NHANES.complete$Months.working <= 60]<- 2 #1-5 years
NHANES.complete$Years.working[NHANES.complete$Months.working > 60 ]<- 3 #More than 5 years

class(NHANES.complete$Years.working) #check the type of variable

NHANES.complete$Years.working<-factor(NHANES.complete$Years.working,labels = c("Less than a year", "1-5 years", "More than 5 years")) #change to factor variable and label levels

#Make schedule into 2 cats
NHANES.complete$Work.Schedule <- car::recode(NHANES.complete$Schedule, 
                              "'9am-5pm'= 'Traditional';
                               'Evenings' = 'Nontraditional';
                               'Early mornings' = 'Nontraditional';
                               'Variable' = 'Nontraditional'")
summary(NHANES.complete)

```

```{r}
#check new cross tabs

#work type vs oral health status
CrossTable(NHANES.complete$Job.Description, NHANES.complete$Oral.health.status)

#years working vs oral health status
CrossTable(NHANES.complete$Years.working, NHANES.complete$Oral.health.status)

#work schedule vs oral health status
CrossTable(NHANES.complete$Work.Schedule, NHANES.complete$Oral.health.status)

#Looks better
```

```{r}
#visualizations

#RQ1 work sector
ggplot(data = NHANES.complete, aes(x = Job.Description)) + 
  geom_bar(aes(fill = Job.Description)) +
  labs(title = "Bar graph of work sector across oral health status") +
  ylab("Frequency") +
  facet_wrap(~Oral.health.status) + 
  scale_fill_manual(values=c("#7fc97f", "#beaed4"), 
                    name = "Work Sector") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

#RQ2 time in job distribution and boxplot
ggplot(data = NHANES.complete, aes(x = Years.working)) + 
  geom_bar(aes(fill = Years.working)) +
  labs(title = "Bar graph of years working in current job across oral health status") +
  ylab("Frequency") +
  facet_wrap(~Oral.health.status) + 
  scale_fill_manual(values=c("#66c2a5", "#fc8d62", "#8da0cb"), 
                    name = "Years Working") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

#RQ3 work schedule
ggplot(data = NHANES.complete, aes(x = Work.Schedule)) + 
  geom_bar(aes(fill = Work.Schedule)) +
  labs(title = "Bar graph of work schedule across oral health status") +
  ylab("Frequency") +
  facet_wrap(~Oral.health.status) + 
  scale_fill_manual(values=c("#a6cee3", "#b2df8a"), 
                    name = "Work Schedule") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

```{r}
#set reference groups

NHANES.complete$Oral.health.status <- relevel(NHANES.complete$Oral.health.status, 
                                            ref = "Good")
NHANES.complete$Race <- relevel(NHANES.complete$Race, 
                                            ref = "Non-Hispanic White")
NHANES.complete$Sex <- relevel(NHANES.complete$Sex, 
                                            ref = "Male")
NHANES.complete$Citizenship <- relevel(NHANES.complete$Citizenship, 
                                            ref = "US Citizen")
NHANES.complete$Insurance <- relevel(NHANES.complete$Insurance, 
                                            ref = "Yes")
NHANES.complete$Work.Schedule <- relevel(NHANES.complete$Work.Schedule, 
                                            ref = "Traditional")
NHANES.complete$Job.Description <- relevel(NHANES.complete$Job.Description, 
                                            ref = "Private")
NHANES.complete$Years.working <- relevel(NHANES.complete$Years.working, 
                                            ref = "Less than a year")
```

```{r}
#Descriptive statistics 
NHANES.complete %>% 
  summarize(mean.age= mean(x=Age),
            sd.age = sd(x=Age),
            med.age=median(x=Age))

```

### Table 1

```{r results = 'markup'}
#table 1
table1(~Age + Race + Citizenship + Sex + Insurance + Job.Description + Work.Schedule + Years.working|Oral.health.status , droplevels=TRUE, data=NHANES.complete)
```


## RQ1 Analysis
### RQ1= Work sector related to oral health

```{r}
#Model with NO covariates
RQ1.mod1 <- multinom(Oral.health.status ~ Job.Description, data=NHANES.complete)
summary(RQ1.mod1)

#hypothesis test
#calculate z score
RQ1.mod1.z <- summary(RQ1.mod1)$coefficients/summary(RQ1.mod1)$standard.errors
# 2-tailed Z test
RQ1.mod1.p <- (1 - pnorm(abs(RQ1.mod1.z), 0, 1))*2
RQ1.mod1.p #give pvalues for each level

# Get OR's
round(exp(coef(RQ1.mod1)), digits = 2)

# Get CI's
round(exp(confint(RQ1.mod1, level=0.95)), digits = 2)
```


```{r}
#model WITH covariates
RQ1.mod2 <- multinom(Oral.health.status ~ Job.Description + Age + Race + Sex + Insurance + Citizenship,
                     data=NHANES.complete)
summary(RQ1.mod2)

#hypothesis test
#calculate z score
RQ1.mod2.z <- summary(RQ1.mod2)$coefficients/summary(RQ1.mod2)$standard.errors
# 2-tailed Z test
RQ1.mod2.p <- (1 - pnorm(abs(RQ1.mod2.z), 0, 1))*2
RQ1.mod2.p #give pvalues for each level

# Get OR's
round(exp(coef(RQ1.mod2)), digits = 2)

# Get CI's
round(exp(confint(RQ1.mod2, level=0.95)), digits = 2)
```

```{r}
#RQ1
#which model predicts better
lrtest(RQ1.mod1, RQ1.mod2)

#model with covariates predicts better
#we will use model with covariates for interpretations
```

#### Results 

```{r results='markup'}
#Table with output of the model with covariates
tab_model(RQ1.mod2, digits = 2, digits.p = 2)

```

## RQ2 Analysis
### RQ2= Consistent income related to oral health

```{r}
#Model with NO covariates
RQ2.mod1 <- multinom(Oral.health.status ~ Years.working, data=NHANES.complete)
summary(RQ2.mod1)

#hypothesis test
#calculate z score
RQ2.mod1.z <- summary(RQ2.mod1)$coefficients/summary(RQ2.mod1)$standard.errors
# 2-tailed Z test
RQ2.mod1.p <- (1 - pnorm(abs(RQ2.mod1.z), 0, 1))*2
RQ2.mod1.p #give pvalues for each level

# Get OR's
round(exp(coef(RQ2.mod1)), digits = 2)

# Get CI's
round(exp(confint(RQ2.mod1, level=0.95)), digits = 2)
```


```{r}
#model WITH covariates
RQ2.mod2 <- multinom(Oral.health.status ~ Years.working + Age + Race + Sex + Insurance + Citizenship, data=NHANES.complete)
summary(RQ2.mod2)

#hypothesis test
#calculate z score
RQ2.mod2.z <- summary(RQ2.mod2)$coefficients/summary(RQ2.mod2)$standard.errors
# 2-tailed Z test
RQ2.mod2.p <- (1 - pnorm(abs(RQ2.mod2.z), 0, 1))*2
RQ2.mod2.p #give pvalues for each level

# Get OR's
round(exp(coef(RQ2.mod2)), digits = 2)

# Get CI's
round(exp(confint(RQ2.mod2, level=0.95)), digits = 2)
```

```{r}
#RQ2
#which model predicts better
lrtest(RQ2.mod1, RQ2.mod2)

#model with covariates predicts better
#we will use model with covariates for interpretations
```

#### Results 

```{r results='markup'}
#Table with output of the model with covariates
tab_model(RQ2.mod2, digits = 2, digits.p = 2)

```

## RQ3 Analysis
### RQ3= Work schedule related to oral health

```{r}
#Model with NO covariates
RQ3.mod1 <- multinom(Oral.health.status ~ Work.Schedule, data=NHANES.complete)
summary(RQ3.mod1)

#hypothesis test
#calculate z score
RQ3.mod1.z <- summary(RQ3.mod1)$coefficients/summary(RQ3.mod1)$standard.errors
# 2-tailed Z test
RQ3.mod1.p <- (1 - pnorm(abs(RQ3.mod1.z), 0, 1))*2
RQ3.mod1.p #give pvalues for each level

# Get OR's
round(exp(coef(RQ3.mod1)), digits = 2)

# Get CI's
round(exp(confint(RQ3.mod1, level=0.95)), digits = 2)
```


```{r}
#model WITH covariates
RQ3.mod2 <- multinom(Oral.health.status ~ Work.Schedule + Age + Race + Sex + Insurance + Citizenship,
                     data=NHANES.complete)
summary(RQ3.mod2)

#hypothesis test
#calculate z score
RQ3.mod2.z <- summary(RQ3.mod2)$coefficients/summary(RQ3.mod2)$standard.errors
# 2-tailed Z test
RQ3.mod2.p <- (1 - pnorm(abs(RQ3.mod2.z), 0, 1))*2
RQ3.mod2.p #give pvalues for each level

# Get OR's
round(exp(coef(RQ3.mod2)), digits = 2)

# Get CI's
round(exp(confint(RQ3.mod2, level=0.95)), digits = 2)
```


```{r}
#RQ3
#which model predicts better
lrtest(RQ3.mod1, RQ3.mod2)

#model with covariates predicts better
#we will use model with covariates for interpretations
```

#### Results 

```{r results='markup'}
#Table with output of the model with covariates
tab_model(RQ3.mod2, digits = 2, digits.p = 2)

```

*Test linearity assumption*

```{r}
#linearity of age

#create term to test linearity
age.linearity <- NHANES.complete$Age * log(NHANES.complete$Age)

#Box Tidwell technique, test the assumption of linearity
boxTidwellAge <- multinom(Oral.health.status ~ Age + age.linearity,
                     data=NHANES.complete) 

summary(boxTidwellAge)

#hypothesis test
#want non-significant interaction term to pass linearity assumption
#calculate z score
boxTidwellAge.z <- summary(boxTidwellAge)$coefficients/summary(boxTidwellAge)$standard.errors
# 2-tailed Z test
boxTidwellAge.p <- (1 - pnorm(abs(boxTidwellAge.z), 0, 1))*2
boxTidwellAge.p #give pvalues for each level

```

- DOES NOT MEET linearity assumption for age

*Check for Multicollinearity*

```{r}
#check for multicollinearity with vif
vif(RQ1.mod2)
vif(RQ2.mod2)
vif(RQ3.mod2)

```

- Age might violate multicollinearity

```{r}
## Project response
# Make a logistic model

#make binary outcome
NHANES.complete$Oral.health <- car:: recode(NHANES.complete$Oral.health.status,
                              "'Poor'= 'Bad';
                               'Moderate' = 'Bad'")
summary(NHANES.complete$Oral.health)

#set new reference 

NHANES.complete$Oral.health <- relevel(NHANES.complete$Oral.health, 
                                            ref = "Good")

summary(NHANES.complete)

```

```{r}
#run logistic regression
logreg <- glm(Oral.health ~ as.factor(Work.Schedule) + Age + as.factor(Race) + as.factor(Sex) + as.factor(Insurance) + as.factor(Citizenship), data=NHANES.complete, family="binomial")
summary(logreg)
  
#calculate and print ORs and 95% CIs  
ORlogreg<-exp(cbind(OR = coef(logreg), confint(logreg))) #calculate ORs and 95% CIs
  ORlogreg #print ORs and 95% CIs
  
tab_model(logreg)
```

